The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-04-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
some improvements in abstract/intro, plus various small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2518 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 248811aa6e84cef6892744882d16aeb2f650775e

\Abstract{
A core step of every algorithm for learning regression trees is the decision if, 
how, and where to split the underlying data - or, in other words, the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employ greedy search strategies 
that directly compare all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selection of covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using $p$-values that are appropriately adjusted for 
the possible split points. In a second step a split point optimizing some objective 
function is selected in the chosen split variable. However, different unbiased tree 
algorithms employ different inference frameworks for computing these $p$-values and 
their relative advantages or disadvantages are not well understood, yet.

Therefore, three different approaches are considered here and embedded into a common 
modeling framework with special emphasis to linear model trees: classical categorical 
association tests (GUIDE), conditional inference (CTree), parameter instability 
tests (MOB). It is assessed how different building blocks affect the power of the 
tree algorithms to select the appropriate covariates for splitting: residuals vs.\ full 
model scores, binarization of residuals/scores at zero, binning of covariates, 
conditional vs.\ unconditional approximations of the null distribution.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with strong variation 
and complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While all of the many developed algorithms 
follow the general idea of splitting the data such that some objective function is optimized, 
they differ in their specific approaches to selecting a split variable and the corresponding 
split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on $p$-values provided by a statistical test. While QUEST, CHAID and GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on asymptotic theory. 
All these approaches have shown to work well for various situations, however, the specific 
advantages and disadvantages of each of the testing strategies have not yet been 
investigated and compared in detail.

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Transformations of the response: using residuals or full model scores
(2)~Binarization of residuals/scores
(3)~Categorization of covariates.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. In that way, the 
effects of each of these features can be investigated separately by varying the building 
blocks in the tool box provided by this general framework. 
Additionally, the impact of applying either a pre- or postpruning strategy is analyzed.

@@

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
After explaining the investigated algorithms CTree (Section~\ref{sec:ctree}), 
MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}) more in detail the applied pruning strategies 
(\ref{sec:pruning}) are discussed.
In Section~\ref{sec:unify} the unifying framework for testing strategies in model-based unbiased 
recursive partitioning algorithms is presented.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are illustrated and discussed in Section~\ref{sec:results}.

\section{Unbiased Recursive Partitioning}
\label{sec:urp}

\subsection{Generic Algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various different tree 
algorithms have been developed, all following the same general structure starting at 
the root of the tree which is simply the first node containing all available data:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached or the data in the node consists of only one or all equal observations. 
\end{enumerate}

@@

\section{Unifying Framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already determined by the testing strategy. Subsequently combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

\begin{table}[t]
\centering
\caption[Table caption text]{Combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.}
\label{tab:combinations}
\begin{tabular}{ l l l l }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & non-parametric  & conditional inference    & pre\\
MOB   & parametric      & score-based fluctuation  & pre (or post with AIC/BIC)\\
GUIDE & parametric      & residual-based $\chi^2$  & post (cost-complexity pruning)\\
\hline
\end{tabular}
\end{table}

\subsection{Unification of Testing Strategies}

@@

\subsection{Building Blocks of Testing Strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
cases.
By including all available scores rather than just residuals the goodness of fit of the model is 
investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

\begin{table}[t]
\centering
\caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the type of test statistic.}
\label{tab:buildingblocks}
\begin{tabular}{ l c c c c }
\hline 
      & Scores       & Binarization  & Categorization  & Statistic\\
\hline
CTree & Model scores & --	           & -- 	           & Sum of squares\\
MOB   & Model scores & --	           & -- 	           & Maximally selected\\
GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
\hline
\end{tabular}
\end{table}

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
hold for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the (arbitrarily chosen) current node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

@@

\section{Simulation Setting and Evaluation}
\label{sec:simulation}

@@

\subsection{Data Generating Process}
Each generated data set consists of a response variable, one regressor, one split 
variable and nine noise variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

\begin{table}
\begin{center}
\caption[Table caption text]{Variables included in the data generating process of 
the first scenario (stump).}
\label{tab:variables}
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

%\newpage
\section{Results}
\label{sec:results}

@@

\subsection{Residuals vs. Full Model Scores}
\label{sec:scores}
To assess the impact of using only residuals or the full score matrix four testing
strategies are compared: CTree, GUIDE, MOB (all three in their original form),
and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_scores).
%Thus, the four strategies are obtained from the unifying framework by setting the switches 
%as listed in Table~\ref{tab:buildingblocks_scores}.

%\begin{table}%[t]
%\centering
%\caption[Table caption text]{CTree, MOB, GUIDE, and GUIDE\_scores with the corresponding setting of
%the switches in the unifying framework and the approximation type of the underlying null 
%distribution.}
%\label{tab:buildingblocks_scores}
%\begin{tabular}{ l c c c c }
%\hline 
%            & Scores            & Binarization  & Categorization  & Null distribution \\ \hline
%CTree       & full model scores & --          & --              & conditional\\
%MOB         & full model scores & --          & --              & unconditional\\
%GUIDE       & residuals         & \checkmark  & \checkmark      & unconditional\\
%GUIDE\_scores & full model scores & \checkmark  & \checkmark      & unconditional\\ \hline
%\end{tabular}
%\end{table}

@@

\subsection{Full Factorial Analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying and the true split point is in the
center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
$\delta = 0.3$ is considered making the differences particularly well observable 
(additional easing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

\begin{table}[h!]
\centering
\caption[Table caption text]{Available levels of the building blocks.}
\label{tab:levels}
\begin{tabular}{ l l l }
\hline
Building Block         & Levels              & Effect \\
\hline
Residuals vs.\ Scores  & \textit{residuals}  & use only residuals\\
                       & \textit{scores}     & use full score matrix\\
Categorization         & \textit{cat}        & categorize split variables\\
                       & \textit{max}        & selected split variables maximally w.o. categorization\\
                       & \textit{lin}        & selected split variables linearly w.o. categorization\\
Binarization           & \textit{bin}        & binarize scores\\
                       & \textit{lin}        & use residuals/score values without binarization\\
\hline
\end{tabular}
\end{table}

@@

\subsection{Prepruning vs. Postpruning}

@@

\section*{Computational Details}
\label{sec:comp}
The applied implementation is based on the \textsf{R} package \textbf{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
where the code to reproduce the simulation study can be found as well.

@@

\section*{Acknowledgements}
\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
2017 to January 2018) was financially supported by the Swiss National
Science Foundation, grant number SNF IZSEZ0 177091.}

@@

\section{Applied Combinations of Building Blocks}\label{app:combinations}
For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``\_res'' refers to using only residuals (instead of all scores),
``\_scores'' to using all scores (instead of only residuals),
``\_cat'' to applying a categorization of split variables,
``\_bin'' to applying a binarization of scores/residuals.

\begin{table}%[t]
\centering
\caption[Table caption text]{Applied testing strategies representing the 12 combinations
of building blocks.}
\label{tab:combinations_appendix}
\begin{tabular}{ l c c c }
\hline 
Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
\hline
GUIDE             & residuals          & bin          & cat \\
MOB\_bin\_res     & residuals          & bin          & max \\
CTree\_bin\_res   & residuals          & bin          & lin \\
MOB\_cat\_res    & residuals          & lin          & cat \\
MOB\_res          & residuals          & lin          & max \\
CTree\_res        & residuals          & lin          & lin \\
GUIDE\_scores     & scores             & bin          & cat \\
MOB\_bin          & scores             & bin          & max \\
CTree\_bin        & scores             & bin          & lin \\
MOB\_cat          & scores             & lin          & cat \\
MOB               & scores             & lin          & max \\
CTree             & scores             & lin          & lin \\
\hline
\end{tabular}
\end{table}

\section{Significance Level}\label{app:significance}
In Section~\ref{sec:scores} the investigated testing strategies are compared based
on the number of cases in which the true split variable $Z_1$ is detected with a $p$-value
smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
is due to the aim of investigating how well the tests perform as significance test.
However, ignoring this significance level yields the same conclusions 
which can be seen when comparing Figures~\ref{fig:scores_stump_wosign}
and \ref{fig:scores_stump}.
In Figure~\ref{fig:scores_stump_wosign} the proportion of cases 
in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
smaller than the significance level, is illustrated. In Figure~\ref{fig:scores_stump} 
the exact same simulation setting is applied with a significance level.

@@

\section{Simulation Results for Continuous Model Parameters}
\label{app:contbeta}
In the simulation study presented in Sections~\ref{sec:simulation} and \ref{sec:results} the intercept
and slope parameters $\beta_0$ and $\beta_1$ are both either fixed or binary variables taking
either the positive or the negative value of the effect size $\delta$. Therefore, these parameters
are step functions for which MOB has shown to perform better than CTree. However, in case of
monotonous functions CTree is advantageous as it is constructed to detect monotonous effects.
To point out that the difference in performance of CTree and MOB depends on the type of effect
the same setting of which the results are presented in Figure~\ref{fig:scores_stump} is
evaluated again but this time with the varying parameter(s) changing continuously.
In particular, the parameters $\beta_0$ and $\beta_1$ are linear functions of the true
split variable $Z_1$:
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{i} \cdot Z_1\\
+\delta \cdot (-1)^{i} \cdot Z_1.
\end{cases}
\end{align*}
Moreover, a modified version of CTree considering maximum values in the test statistic is evaluated
(denoted by CTree\_max). 
Looking at the results presented in Figure~\ref{fig:stump_contbeta} it can be observed that in this setting the 
original version of CTree is ahead while CTree\_max and MOB perform almost equally well. Hence, the comparison of 
Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} points out that it clearly depends on the type of 
effect whether the maximal selection (MOB, CTree\_max) or the linear selection (CTree) is advantageous.

@@

\section{Results Over Increasing Effect Size}
\label{app:increasingdelta}

>>>>

\Abstract{
A core step of every algorithm for learning regression trees is the decision if, 
how, and where to split the underlying data - or, in other words, the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employed greedy search strategies 
that directly compare all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selection of covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using $p$-values that are appropriately adjusted for 
the possible split points. In a second step a split point optimizing some objective 
function is selected in the chosen split variable. However, different unbiased tree 
algorithms employ different inference frameworks for computing these $p$-values and 
their relative advantages or disadvantages are not well understood, yet.

Therefore, three different popular approaches are considered here: classical categorical 
association tests (GUIDE), conditional inference (CTree), and parameter instability 
tests (MOB). First, these are embedded into a common inference framework encompassing
parametric model trees, in particular linear model trees. Second, it is assessed how
different building blocks from this common framework affect the power of the 
tree algorithms to select the appropriate covariates for splitting: observation-wise
goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
at zero, binning of covariates, conditional vs.\ unconditional approximations of the
null distribution. This shows that specifically the goodness-of-fit measure is crucial
for the power of the procedures, with model scores without binarization performing
much better in many scenarios.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with strong variation 
and complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While almost all algorithms proposed in the literature
follow the general idea of splitting the data such that some objective function is optimized, 
they differ in their specific approaches to selecting a split variable and the corresponding 
split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on $p$-values provided by a statistical test. While QUEST, CHAID, and GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on central limit theorems
for the parameter estimators. 
All these approaches have shown to work well for various situations, however, the relative
(dis)advantages of the testing strategies have not yet been investigated and compared in detail.

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Transformations of a goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Binarization of these residuals or scores.
(3)~Categorization of covariates.
(4)~Approximations of the null distribution: conditional vs.\ unconditional.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

The remainder of this manuscript is structured as follows:
Section~\ref{sec:urp} reviews unbiased tree models, starting with the general algorithmic idea
(Section~\ref{sec:generic_alg}) followed by the specific algorithms CTree (Section~\ref{sec:ctree}), 
MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}), before possible pruning strategies 
(\ref{sec:pruning}) are discussed.
In Section~\ref{sec:unify} the unifying framework for testing strategies in model-based unbiased 
recursive partitioning algorithms is presented.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are illustrated and discussed in Section~\ref{sec:results}.

\section{Unbiased recursive partitioning}
\label{sec:urp}

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various different tree 
algorithms have been developed, all following the same general structure starting at 
the root of the tree which is simply the first node containing all available data:
\begin{enumerate}
\item For the data set in the current node some form of discrepancy 
measure is calculated. 
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached. 
\end{enumerate}

@@

\section{Unifying framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already determined by the testing strategy. Subsequently combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & Non-parametric  & Conditional inference    & Pre\\
MOB   & Parametric      & Score-based fluctuation  & Pre (or post with AIC/BIC)\\
GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
\hline
\end{tabular}
\caption{Combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.
\label{tab:combinations}}
\end{table}

\subsection{Unification of testing strategies}

@@

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
cases.
By including all available scores rather than just residuals the goodness of fit of the model is 
investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l l }
\hline 
      & Scores       & Binarization  & Categorization  & Statistic\\
\hline
CTree & Model scores & --	     & -- 	       & Sum of squares\\
MOB   & Model scores & --	     & -- 	       & Maximally selected\\
GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
\hline
\end{tabular}
\caption{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the type of test statistic.
\label{tab:buildingblocks}}
\end{table}

@@

\subsection{Linear model tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
hold for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the (arbitrarily chosen) current node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

@@

\section{Simulation setting and evaluation}
\label{sec:simulation}

@@

\subsection{Data generating process}
Each generated data set consists of a response variable, one regressor, one split 
variable and nine noise variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

\begin{table}
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{Variables included in the data generating process of 
the first scenario (stump).
\label{tab:variables}}
\end{table}

@@

\section{Results}
\label{sec:results}

@@

\subsection[Residuals vs. full model scores]{Residuals vs.\ full model scores}
\label{sec:scores}
To assess the impact of using only residuals or the full score matrix four testing
strategies are compared: CTree, GUIDE, MOB (all three in their original form),
and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_scores).

@@

\subsection{Full factorial analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying and the true split point is in the
center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
$\delta = 0.3$ is considered making the differences particularly well observable 
(additional easing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building Block         & Levels              & Effect \\
\hline
Residuals vs.\ scores  & \textit{residuals}  & Use only residuals\\
                       & \textit{scores}     & Use full score matrix\\
Categorization         & \textit{cat}        & Categorize split variables\\
                       & \textit{max}        & Selected split variables maximally w.o. categorization\\
                       & \textit{lin}        & Selected split variables linearly w.o. categorization\\
Binarization           & \textit{bin}        & Binarize scores\\
                       & \textit{lin}        & Use residuals/score values without binarization\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\subsection[Pre-pruning vs. post-pruning]{Pre-pruning vs.\ post-pruning}

@@

\section*{Computational details}
\label{sec:comp}
The applied implementation is based on the \textsf{R} package \textbf{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
where the code to reproduce the simulation study can be found as well.

@@

\section*{Acknowledgments}
\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
2017 to January 2018) was financially supported by the Swiss National
Science Foundation, grant number SNF IZSEZ0 177091.}

@@

\section{Combinations of building blocks applied}\label{app:combinations}
For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``\_res'' refers to using only residuals (instead of all scores),
``\_scores'' to using all scores (instead of only residuals),
``\_cat'' to applying a categorization of split variables,
``\_bin'' to applying a binarization of scores/residuals.

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline 
Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
\hline
GUIDE             & residuals          & bin          & cat \\
MOB\_bin\_res     & residuals          & bin          & max \\
CTree\_bin\_res   & residuals          & bin          & lin \\
MOB\_cat\_res    & residuals          & lin          & cat \\
MOB\_res          & residuals          & lin          & max \\
CTree\_res        & residuals          & lin          & lin \\
GUIDE\_scores     & scores             & bin          & cat \\
MOB\_bin          & scores             & bin          & max \\
CTree\_bin        & scores             & bin          & lin \\
MOB\_cat          & scores             & lin          & cat \\
MOB               & scores             & lin          & max \\
CTree             & scores             & lin          & lin \\
\hline
\end{tabular}
\caption{Testing strategies applied (representing the 12 combinations
of building blocks).
\label{tab:combinations_appendix}}
\end{table}

\section{Significance level}\label{app:significance}
In Section~\ref{sec:scores} the investigated testing strategies are compared based
on the number of cases in which the true split variable $Z_1$ is detected with a $p$-value
smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
is due to the aim of investigating how well the tests perform as significance test.
However, ignoring this significance level yields the same conclusions 
which can be seen when comparing Figures~\ref{fig:scores_stump_wosign}
and \ref{fig:scores_stump}.
In Figure~\ref{fig:scores_stump_wosign} the proportion of cases 
in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
smaller than the significance level, is illustrated. In Figure~\ref{fig:scores_stump} 
the exact same simulation setting is applied with a significance level.

@@

\section{Simulation results for continuous model parameters}
\label{app:contbeta}
In the simulation study presented in Sections~\ref{sec:simulation} and \ref{sec:results} the intercept
and slope parameters $\beta_0$ and $\beta_1$ are both either fixed or binary variables taking
either the positive or the negative value of the effect size $\delta$. Therefore, these parameters
are step functions for which MOB has shown to perform better than CTree. However, in case of
monotonous functions CTree is advantageous as it is constructed to detect monotonous effects.
To point out that the difference in performance of CTree and MOB depends on the type of effect
the same setting of which the results are presented in Figure~\ref{fig:scores_stump} is
evaluated again but this time with the varying parameter(s) changing continuously.
In particular, the parameters $\beta_0$ and $\beta_1$ are linear functions of the true
split variable $Z_1$:
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{i} \cdot Z_1\\
+\delta \cdot (-1)^{i} \cdot Z_1.
\end{cases}
\end{align*}
Moreover, a modified version of CTree considering maximum values in the test statistic is evaluated
(denoted by CTree\_max). 
Looking at the results presented in Figure~\ref{fig:stump_contbeta} it can be observed that in this setting the 
original version of CTree is ahead while CTree\_max and MOB perform almost equally well. Hence, the comparison of 
Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} points out that it clearly depends on the type of 
effect whether the maximal selection (MOB, CTree\_max) or the linear selection (CTree) is advantageous.

@@

\section{Results for increasing effect size}
\label{app:increasingdelta}
